Icicle: Write Once, Run Once

### 2016-07-26 2016fhpc/section/02-Source.tex
Some notes from reviewers

Here we use the @group@ construct to assign key-value pairs to groups as we obtain them, and compute the running mean of the values of each group. The @avgs@ value becomes a map of group keys to their running means. Once we reach the end of the stream we will have access to the last key and can lookup the final result.

@@

In Icicle, queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment. In Icicle, the stream being processed (such as @xs@ above) is implicit in each query. Constructs such as @filter@ and @fold@ do not take the name of the stream as an argument, but instead operate on the stream defined in the context. Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query. In the expression @(filter p of mean value)@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.

@@

The grammar for Icicle is given in Figure~\ref{fig:source:grammar}.
Value types $T$ include numbers, booleans and maps.
Modal types $\TauMode$ include the pure value types, and modalities associated with a value type.
Function types $\TauFun$ include non-function modal types, and functions from modal type to function type.
As Icicle is a first-order language, function types are not value types.

>>>>

Here we use the @group@ construct to assign key-value pairs to groups as we obtain them, and compute the running mean of the values of each group. The @avgs@ value becomes a map of group keys to their running means. Once we reach the end of the stream we will have access to the last key and can lookup the final result.
%% Review #1:
%% In Section 3 I kept looking for a definition of the “last” function, which takes an important role in the Introduction.
%% In Section 4, I found the definition. It would help the reader to give a small note early in Section 2 or 3 that “last”, “mean”, “sum” are user-level (i.e. non-primitive) functions whose definition will be given in Section 4.
Evaluation and typing rules are defined in~\S\ref{s:IcicleSource}, while the user functions @last@ and @mean@ are defined in~\S\ref{s:IcicleCore}.

@@

%% Review #1
%% It’s not clear what’s meant by “In Icicle, queries that would require unbounded buffering cannot be written”.
%% My understanding is that they can be written, but we would get a runtime error because the buffer would eventually become too big to fit the memory. Is this right?
% change to "statically outlawed by the typesystem": I think that's unambiguous.

In Icicle, queries that would require unbounded buffering are statically outlawed by the typesystem, with one major caveat that we will discuss in a moment. In Icicle, the stream being processed (such as @xs@ above) is implicit in each query. Constructs such as @filter@ and @fold@ do not take the name of the stream as an argument, but instead operate on the stream defined in the context. Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query. In the expression @(filter p of mean value)@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.

@@

The grammar for Icicle is given in Figure~\ref{fig:source:grammar}.
Value types $T$ include numbers, booleans and maps.
Modal types $\TauMode$ include the pure value types, and modalities associated with a value type.
%% Review 3:
%% In Section 3, the description of function types is confusing.
%% The authors state, "Function types F include non-function modal types, and functions from modal type to function type."
%% It appears from Figure 1 that function types do not include non-function modal types (syntactic category M), and include only functions from modal type(s) to *modal* type.
Function types $\TauFun$ include functions with any number of modal type arguments to a modal return type.
As Icicle is a first-order language, function types are not value types.

### 2016-07-26 2016fhpc/section/04-Benchmarks.tex
Some notes from reviewers

In our proof of concept tesing we replaced an existing R script that performed feature generation with new Icicle code. The R script computed features from a 317GB data set. It computed 12 queries over each of 31 input tables, for 372 query evaluations in total. The R script took 15 hours to run and consisted of 3,566 lines of code. The replacement Icicle version is only 191 lines of code and takes seven minutes to run.

@@

We ran all the Unix utilities with unicode decoding disabled using @LANG=C LC_COLLATE@ for maximum performance. The input data does not contain unicode characters. 
We used an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage. The fused Icicle version significantly outperformed the R version of the queries, and the single-threaded version was on par with @wc@, while only a little slower than @grep@. This is despite the fact that the Icicle queries perform more computational work than than @wc@ and @grep@. By using multiple processors, we were able to scale up to perform as well as @cat@, approaching the disk speed.
The memory usage of Icicle starts at around 200MB of RAM for a single thread, but as more threads are added approaches 15MB per thread.
The memory usage is constant in the input size and depends on the number of queries.
The R code is single threaded and would require at least 150 processors to reach similar speeds, assuming perfect scaling.
These results give us confidence that our distributed implementation will be fast as well as scalable~\cite{mcsherry2015scalability}.

>>>>

In our proof of concept testing we replaced an existing R script that performed feature generation with new Icicle code. The R script computed features from a 317GB data set. It computed 12 queries over each of 31 input tables, for 372 query evaluations in total. The R script took 15 hours to run and consisted of 3,566 lines of code. The replacement Icicle version is only 191 lines of code and takes seven minutes to run.

@@

We ran all the Unix utilities with unicode decoding disabled using @LANG=C LC_COLLATE@ for maximum performance. The input data does not contain unicode characters. 
We used an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage. The fused Icicle version significantly outperformed the R version of the queries, and the single-threaded version was on par with @wc@, while only a little slower than @grep@. This is despite the fact that the Icicle queries perform more computational work than @wc@ and @grep@. By using multiple processors, we were able to scale up to perform as well as @cat@, approaching the disk speed.
The memory usage of Icicle starts at around 200MB of RAM for a single thread, but as more threads are added approaches 15MB per thread.
The memory usage is constant in the input size and depends on the number of queries.
The R code is single threaded and would require at least 150 processors to reach similar speeds, assuming perfect scaling.
These results give us confidence that our distributed implementation will be fast as well as scalable~\cite{mcsherry2015scalability}.

### 2016-07-26 2016fhpc/section/XX-Related.tex
Some notes from reviewers

StreamIt~\cite{thies2002streamit} is an imperative streaming language which has been extended with dynamic scheduling~\cite{soule2013dynamic}. Dynamic scheduling handles data flow graphs where the transfer rate between different stream operators is not known at compile time. Dynamic scheduling is trade-off: it is required for stream operators such as grouping and filtering where the output data rate is not known statically, but using dynamic techniques for graphs with static transfer rates tends to have a performance cost. Icicle includes grouping and filtering operators where the output rates are statically unknown, however the associated language constructs require grouped and filtered data to be aggregated rather than passed as as the input to another stream operator. This allows Icicle to retain fully static scheduling, so the compiled queries consist of straight line code with no buffering.

>>>>

StreamIt~\cite{thies2002streamit} is an imperative streaming language which has been extended with dynamic scheduling~\cite{soule2013dynamic}. Dynamic scheduling handles data flow graphs where the transfer rate between different stream operators is not known at compile time. Dynamic scheduling is trade-off: it is required for stream operators such as grouping and filtering where the output data rate is not known statically, but using dynamic techniques for graphs with static transfer rates tends to have a performance cost. Icicle includes grouping and filtering operators where the output rates are statically unknown, however the associated language constructs require grouped and filtered data to be aggregated rather than passed as the input to another stream operator. This allows Icicle to retain fully static scheduling, so the compiled queries consist of straight line code with no buffering.
